Appendix 5: Comparison of index standardization of fishery-independent survey data with VAST and sdmTMB

### 2022-01-28 doc/appendix-pcod-VAST-tweedie.Rmd
Copy edit/ slightly jitter VAST

We will fit a geostatistical spatiotemporal model with VAST for the purposes of index standardization, following the sdmTMB example. We will use a data set built into the sdmTMB package: trawl survey data for Pacific Cod in Queen Charlotte Sound. The density units are kg/km^2^. Here, X and Y are coordinates in UTM zone 9.

To specify the mesh used to approximate the spatial process, which is used in the SPDE calculations, we use the k-means method in VAST. Rather than specifying the cutoff distance, meshes in VAST are typically generated by specifying only the number of knots, which we will later pass, along with other model settings to the function make_settings. We will use 205 knots, the same number in the mesh created in the other versions of this appendix.

For this example, we will include a factor predictor that represents the mean estimate for each time slice. Settings used for index standardization are specified partially by defining `purpose = "index2"` but we also explicitly provide arguments for these and other key settings here. Specifically, we set the elements of `RhoConfig` to 0 to specify that we want the intercepts and spatiotemporal fields to be independent (which is the default).

VAST is constructed to accommodate many different model structures, including delta or hurdle models, and thus requires the user to specify arguments for the spatial and spatiotemporal fields of two model components. In this case, we want to fit a model with a single component, using the Tweedie distribution. To do so, we must turn off estimation of omega and epsilon for 1st linear predictor by entering 0 for those elements in the object we'll use to define the configuration of these fields `FieldConfig`.

@@

Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function 'fit_model' and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST). Several options are built in for commonly used regions, but in this example we will need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in `make_setting()` and `input_grid = input_grid` in `fit_model()`.

@@

We can compare the various parameter estimates. Note that VAST adds the dispersion parameter (`phi` in sdmTMB) in the mean of the Tweedie distribution, so we need to account for that when comparing main-effect estimates.

@@

Now, we can compare the indices.

@@

sdm_i <- ind_sdmTMB %>% mutate(index = "sdmTMB")
both_i <- bind_rows(sdm_i, vast_i) %>% filter(est > 0)
ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
  geom_ribbon(alpha = 0.3) +
  geom_line() + 
  ylim(0, max(both_i$upr)) +
  coord_cartesian(expand = FALSE) +
  ylab("Biomass (t)")

@@

The results are effectively identical. In the ribbon plot above, the lines perfectly cover each other.

>>>>

We will fit a geostatistical spatiotemporal model with VAST for the purposes of index standardization. We will use a data set built into the sdmTMB package: trawl survey data for Pacific Cod in Queen Charlotte Sound. The density units are kg/km^2^. Here, X and Y are coordinates in UTM zone 9.

To specify the mesh used to approximate the spatial process, which is used in the SPDE calculations, we use the k-means method in VAST. Rather than specifying the cutoff distance, meshes in VAST are typically generated by specifying only the number of knots, which we will later pass, along with other model settings to the function `make_settings()`. We will use 205 knots, the same number in the mesh created in the other appendices.

For this example, we will include a factor predictor that represents the mean estimate for each time slice. Settings used for index standardization are specified partially by defining `purpose = "index2"` but we also explicitly provide arguments for these and other key settings here. Specifically, we set the elements of `RhoConfig` to `0` to specify that we want the intercepts and spatiotemporal fields to be independent, which is the default.

VAST is constructed to accommodate many different model structures, including delta or hurdle models, and thus requires the user to specify arguments for the spatial and spatiotemporal fields of two model components. In this case, we want to fit a model with a single component, using the Tweedie distribution. To do so, we must turn off estimation of omega and epsilon for first linear predictor by entering `0` for those elements in the object we will use to define the configuration of these fields in `FieldConfig`.

@@

Unlike in sdmTMB, the fitting and predicting steps are all accomplished with the function `fit_model()` and thus we need to specify the prediction grid (referred to as the "extrapolation grid" in VAST). Several options are built in for commonly used regions, but in this example we will need to format the Queen Charlotte Sound grid in sdmTMB to be used in VAST and specify `Region = "User"` in `make_setting()` and `input_grid = input_grid` in `fit_model()`.

@@

We can compare the various parameter estimates. Note that VAST adds the dispersion parameter (`phi` in sdmTMB) into the mean of the Tweedie distribution, so we need to account for that when comparing main-effect estimates.

@@

Now, we can compare the indices. In the ribbon plot below, the lines perfectly covered each other and so we added a small jitter amount to make them both visible.

@@

# jitter slightly for visualization:
jitter_amount <- mean(vast_i$est) * 0.01
vast_i_gg <- vast_i
vast_i_gg$est <- vast_i_gg$est + jitter_amount
vast_i_gg$lwr <- vast_i_gg$lwr + jitter_amount
vast_i_gg$upr <- vast_i_gg$upr + jitter_amount

sdm_i <- ind_sdmTMB %>% mutate(index = "sdmTMB")
both_i <- bind_rows(sdm_i, vast_i_gg) %>% filter(est > 0)
ggplot(both_i, aes(x = year, y = est, ymin = lwr, ymax = upr, colour = index)) + 
  geom_ribbon(fill = NA, lty = 2) +
  geom_line() + 
  ylim(0, max(both_i$upr)) +
  coord_cartesian(expand = FALSE) +
  ylab("Biomass (t)") +
  theme_light()

@@

The results are effectively identical.
